Dish: Distribution-oblivious Shell Scripting

### 2019-11-22 pldi/draft.tex
Updates on commands

\newcommand{\pur}{\cn{\textsc{P}}}
\newcommand{\sta}{\cn{\textsc{S}}}
\newcommand{\dfs}{\cn{\textsc{F}}}
\newcommand{\sid}{\cn{\textsc{E}}}
\newcommand{\irr}{\cn{\textsc{I}}}

\newcommand{\tpur}{\tcn{\textsc{P}}}
\newcommand{\tsta}{\tcn{\textsc{S}}}
\newcommand{\tdfs}{\tcn{\textsc{F}}}
\newcommand{\tsid}{\tcn{\textsc{E}}}
\newcommand{\tirr}{\tcn{\textsc{I}}}

@@

%% Title information
\title{Distribution-oblivious Programming with Dish}         %% [Short Title] is optional;
% \titlenote{with title note}             %% \titlenote is optional;
%                                         %% can be repeated if necessary;
%                                         %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

\kk{The title is too abstract and doesn't really hint that we
  distribute shell code. Also we could consider capitalizing the name
  as DiSh?}

@@

To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of individual commands involved in them.
This understanding is encoded in a command's \emph{distributability} class.
Put informally, a distributability class representing the level of synchronization required by copies of a command executing concurrently.
The class captures characteristics that are imprortant for distribution, rather than all the observable behavior of a command.
\sys leans towards a few, coarse classes rather than many, detailed ones---after all, these classes should be easily understood by command developers.
% Make the connections that this is like a type, that doesn't capture everything, but is good-enough

The section starts with a distributability study of standard libraries commands, such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
Building on this study, it presents a DSL for describing a command's distributability characteristics~\sx{ext}.
In return, \sys uses this DSL to tag commands in the standard libraries and generate the transformations presented in later sections.

@@

% nv TODO: this is a hierarchy
Broadly, shell commands can be split into five major classes with respect to their distribution characteristics, depending on how they interact with state.
Tab.~\ref{tab:classes} summarizes the discussion;
These classes are listed (and presented later) in ascending difficulty of distribution; later classes require further effort.
In this order, each class can be thought as a subset of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).

\begin{table}[t]
\center
\footnotesize
\setlength\tabcolsep{3pt}
\caption{
  \footnotesize{
    \textbf{Distributability Classes}.
    Broadly, \unix commands can be broken down into to five classes.
  }
}
\begin{tabular}{l @{\extracolsep{\fill}} llll}
\toprule
Class                           &  Key    & Example Commands                            & Coreutils       & POSIX       \\
\midrule
Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)     &           \\  % 22
Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)     &           \\  % 21
(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)     &           \\  % 27
Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)       &           \\  % 25
Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{restart}     &  5  (0.4\%)      &           \\  % 5
% \midrule
% Shell PL Constructs             &                         &                  \\
% \etc                            &                         &                  \\
\bottomrule
\end{tabular}
\label{tab:classes}
\end{table}

@@

\heading{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands expressible as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness,
  they do not need care planning on where to split.
 % map :: (a -> b) -> [a] -> [b]

Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time);
  this could potentially allow further parallelization within a single line element.
However, exactly because of their local, commands operating on individual characters are not particularly expensive computationally.

\heading{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity, returning the same results for the same inputs;
  however, they need to maintain internal state across the entire invocation.

Pure commands often cannot operate in an online fashion, but need to block until the end of the stream.
A typical example of this is \ttt{sort}, a command that cannot start before the last line of input has been consumed.

Parallelizing these commands is slightly more challenging and depends on the command.
Some commands have known divide-and-conquer equivalents.

Pure commands~ones that maintain 
Distribution:

\heading{File-system Commands}

This is the largest class, as the file (and file-system) abstraction is so central to \unix.
In fact, \unix proxies several file-unrelated operators through the file-system---for example, security in \unix is grafted upon the file-system:

(and all associated commands) are always
And most of these commands seem control-plane operations that operate on identifiers~\sx{bg}---for example, renaming a file or changing group ownership.
% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
These are mostly distributable

Distribution:

\heading{Side-effectful Commands}
Another large class of commands are ones that have side-effects to memory.
Because of timesharing and multi-tasking (many users can operate multiple terminals on single \unix machine), a set of commands are devoted to accessing variables---environment variables, shell parsing flags, 
The different formats have multiplied these commands

Thankfully, the vast majority of these commands (a ratio of 21:4) only \emph{read} values;
Even better,  usually read values that are not altered by user---\eg \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information.
Even if \sid commands are protected with transactions
these benefits are even greater for read-only or read-mostly accesses.

\heading{Irreversibly Side-effectful Commands}

>>>>

\newcommand{\pur}{\cn{\textsc{P}}\xspace}
\newcommand{\sta}{\cn{\textsc{S}}\xspace}
\newcommand{\dfs}{\cn{\textsc{F}}\xspace}
\newcommand{\sid}{\cn{\textsc{E}}\xspace}
\newcommand{\irr}{\cn{\textsc{I}}\xspace}

\newcommand{\tpur}{\tcn{\textsc{P}}\xspace}
\newcommand{\tsta}{\tcn{\textsc{S}}\xspace}
\newcommand{\tdfs}{\tcn{\textsc{F}}\xspace}
\newcommand{\tsid}{\tcn{\textsc{E}}\xspace}
\newcommand{\tirr}{\tcn{\textsc{I}}\xspace}

@@

%% Title information
\title{Dish: Distribution-oblivious Shell Scripting}         %% [Short Title] is optional;
% \titlenote{with title note}             %% \titlenote is optional;
%                                         %% can be repeated if necessary;
%                                         %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

@@

To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of their intermediate commands.
This understanding is encoded in a command's \emph{distributability} class.
Informally, a distributability class represents the level of synchronization required by copies of a command executing in parallel.
It focuses on characteristics that are important for distribution, rather than all the observable behavior of a command.
\sys leans towards a few coarse classes rather than many detailed ones---among other reasons, to simplify their use by command developers.
% Make the connections that this is like a type, that doesn't capture everything, but is good-enough

This section starts by defining these classes, paired with a distributability study of the commands found in standard libraries such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
Building on this study, it presents a DSL that enables command classification by its developers~\sx{ext}.
In return, \sys uses this DSL to tag commands in the standard libraries and generate their wrappers presented in later sections.

@@

% nv TODO: this is a hierarchy
Broadly, shell commands can be split into five major classes (summarized in Tab.~\ref{tab:classes}) with respect to their distribution characteristics, depending on how they interact with state.
These classes are ordered in ascending difficulty of distribution:
  later classes require more effort.
In this order, each class can be thought as a subset of the next one---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).

\begin{table}[t]
\center
\footnotesize
\setlength\tabcolsep{3pt}
\caption{
  \footnotesize{
    \textbf{Distributability Classes}.
    Broadly, \unix commands can be broken down into to five classes.
  }
}
\begin{tabular}{l @{\extracolsep{\fill}} llll}
\toprule
Class                           &  Key    & Example Commands                            & Coreutils       & POSIX       \\
\midrule
Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)     &           \\  % 22
Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)     &           \\  % 21
(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)     &           \\  % 27
Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)       &           \\  % 25
Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)      &           \\  % 5
% \midrule
% Shell PL Constructs             &                         &                  \\
% \etc                            &                         &                  \\
\bottomrule
\end{tabular}
\label{tab:classes}
\end{table}

@@

\heading{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands that can be expressed as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to split.
 % map :: (a -> b) -> [a] -> [b]

Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time).
This property could allow further parallelization within a single line element, a feature that might seem of limited use because these commands are not particularly expensive computationally precisely due to their narrow focus.
However, it is useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.

\heading{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity, returning the same results for the same inputs, but maintain internal state across their entire pass.
The details of this state and its propagation during piecewise element processing affects their distributability characteristics.

Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc}, which simply maintains a counter.
Others, such as \ttt{sha1sum}, maintain more complex state that is not applied commutatively.

Often these commands do not operate in an online fashion, but need to block until the end of a stream---a typical example of this is \ttt{sort}, which cannot start before the last line of input has been consumed.
Non-streaming constraints affect task parallelism, but not data parallelism:
  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques.

\heading{File-system Commands}
The third class contains commands that access the file-system~\dfs.
This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving is grafted upon the file-system.

% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
Commands in \dfs are distributable for many reasons, as evidenced by a long history of distributed \unix file-system clones~\cite{catalogue}.
Some of these offer the semantics of the \unix file-system---hierarchies, protection bits---and a POSIX virtualization layer that enables support for the commands in \dfs.
The vast majority of these commands are control-plane operations that act on identifiers~\sx{bg}---for example, \ttt{touch} for creating a file or \ttt{chgrp} for changing group ownership.
% these operate by changing an inode
Data-plane operations on files go explicitly through stream-redirection operations~\sx{bg}.
Operations using different identifiers pose no problems;
  operations sharing a single identifier might have problems---but these risks exist even in the non-distributed case.
While such pathological cases are possible, they are rare enough---and have been rare enough since \unix's inception~\cite[\S3.6]{unix}).

% \unix itself does not offer any guarantees when a file is accessed concurrently on a single computer.

\heading{Side-effectful Commands}
Another large class of commands are ones that have side-effects to memory~\sid.
Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.

A command in \sid will require some form of concurrency control to distribute property.
However, because its side-effects stay within the distributed \unix environment, they can be manipulated using---including optimistic and multi-versioning schemes that can roll-back side-effects.

The vast majority of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never affected by the user.
For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
Even if \sid commands are protected with transactions
  these benefits are even greater for read-only or read-mostly accesses.

\heading{Irreversibly Side-effectful Commands}
Finally, some commands have side-effects outside \sys---for example, by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
Such commands are not distributable---they have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism.
